taxadb: A High-Performance Local Taxonomic Database Interface

### 2019-12-10 paper/manuscript.Rmd
Citations by hand in the provider table.

<!--

Table version of below - citation, number of unique identifier

- the Integrated Taxonomic Information System [ITIS; @itis], original formed to standardize taxonomic name usage across many agencies in the United States federal government, 
- the National Center for Biological Information's (NCBI) Taxonomy database, [@ncbi]
- Catalogue of Life (COL) Annual Species list. (2018 version at present)
- Global Biodiversity Information Facility Taxonomic Backbone [GBIF; @gbif].  
- FishBase [@fishbase] taxonomic names
- SeaLifeBase [@sealifebase] taxonomic names
- Wikidata [@jorrit]
- Open Tree Taxonomy, OTT
- International Union for Conservation of Nature and Natural Resources (IUCN) Red list @iucn

(ref:itis) Integrated Taxonomic Information System [@itis]
-->

```{r echo = FALSE, results = "asis"}
# #provider descriptions
desc <- c(itis = "originally formed to standardize taxonomic name usage across many agencies in the United States federal government",
          ncbi = "nomenclature for sequences in the International Nucleotide Sequence Database Collaboration database",
          col = "comphrehensive taxonomic effort, includes some other providers (e.g. itis)",
          gbif = "taxonomic backbone of the GBIF database, assembled from other sources including COL",
          fb = "nomenclature for global database of fishes",
          wd = "",
          ott = "comphrenesive tree of life based on phylogenetic trees and taxonomic data",
          iucn = "taxonomy for classification of species status")

tibble(provider = c("Integrated Taxonomic Information System, (ref:@itis)",
                                 "National Center for Biological Information's Taxonomy database, @ncbi",
                                 "Catalogue of Life, @col",
                                 "Global Biodiversity Information Facility Taxonomic Backbone, @gbif",
                                 "FishBase, @fishbase",
                                 #"SeaLifeBase",
                                 "Wikidata, @jorrit", "Open Tree Taxonomy, @Rees2017",
                                 "International Union for Conservation of Nature and Natural Resources, @iucn"),
                    abbreviation = c("itis", "ncbi", "col", "gbif", "fb", #"slb",
                                     "wd", "ott", "iucn"),
                    total_identifiers = map(abbreviation,
                                            ~taxa_tbl(.x) %>%
                                              select(acceptedNameUsageID) %>% pull() %>% n_distinct()),
                    description = desc
) %>%
  knitr::kable(format = "latex", col.names = c("Provider", "Abbreviation", "Number of \nIdentifiers", "Description"), escape = FALSE) %>%
  kable_styling(latex_options="scale_down") %>%
  kableExtra::column_spec(1, width = "5cm") %>%
  kableExtra::column_spec(4, width = "10cm")

@@

Taxonomic indentifiers are a powerful tool for maintaing taxonomic consistency, an essential task for a wide variety of applications. Despite multiple taxonomic standardization efforts, resolving names to taxonomic identifiers is often not a standard step in the research work flow due to difficulty in accessing providers and the time consuming API queries necessary for resolving even moderately sized data sets. `taxadb` fills an important gap between existing tools and typical research patterns by providing a fast, reproducible approach for matching names to taxonomic identifiers. `taxadb` is not intended as an improvement or replacement for any existing approaches to taxonomic name resolution. In particular, `taxadb` is not a replacement for the APIs or databases provided, but merely an interface to taxonomic naming information contained within that data. 
----

>>>>

<!--
Table version of below - citation, number of unique identifier

- the Integrated Taxonomic Information System [ITIS; @itis], original formed to standardize taxonomic name usage across many agencies in the United States federal government, 
- the National Center for Biological Information's (NCBI) Taxonomy database, [@ncbi]
- Catalogue of Life (COL) Annual Species list. (2018 version at present)
- Global Biodiversity Information Facility Taxonomic Backbone [GBIF; @gbif].  
- FishBase [@fishbase] taxonomic names
- SeaLifeBase [@sealifebase] taxonomic names
- Wikidata [@jorrit]
- Open Tree Taxonomy, OTT
- International Union for Conservation of Nature and Natural Resources (IUCN) Red list @iucn
-->

```{r echo = FALSE, results = "asis"}
#provider descriptions
desc <- c(itis = "originally formed to standardize taxonomic name usage across many agencies in the United States federal government",
          ncbi = "nomenclature for sequences in the International Nucleotide Sequence Database Collaboration database",
          col = "comphrehensive taxonomic effort, includes some other providers (e.g. itis)",
          gbif = "taxonomic backbone of the GBIF database, assembled from other sources including COL",
          fb = "nomenclature for global database of fishes",
          ott = "comphrenesive tree of life based on phylogenetic trees and taxonomic data",
          iucn = "taxonomy for classification of species status")

tibble(provider = c("Integrated Taxonomic Information System, ITIS 2019",
                                 "National Center for Biological Information's Taxonomy database, NCBI",
                                 "Catalogue of Life, COL",
                                 "Global Biodiversity Information Facility Taxonomic Backbone, GBIF 2019",
                                 "FishBase, Froese and Pauly 2019",
                                 "Open Tree Taxonomy, OTT",
                                 "International Union for Conservation of Nature and Natural Resources,IUCN 2019"),
                    abbreviation = c("itis", "ncbi", "col", "gbif", "fb", "ott", "iucn"),
                    total_identifiers = map(abbreviation,
                                            ~taxa_tbl(.x) %>%
                                              select(acceptedNameUsageID) %>% pull() %>% n_distinct()),
                    description = desc
) %>%
  knitr::kable(format = "latex", col.names = c("Provider", "Abbreviation", "Number of \nIdentifiers", "Description"), escape = FALSE) %>%
  kable_styling(latex_options="scale_down") %>%
  kableExtra::column_spec(1, width = "5cm") %>%
  kableExtra::column_spec(4, width = "10cm")

@@

Taxonomic indentifiers are a powerful tool for maintaing taxonomic consistency, an essential task for a wide variety of applications. Despite multiple taxonomic standardization efforts, resolving names to taxonomic identifiers is often not a standard step in the research work flow due to difficulty in accessing providers and the time consuming API queries necessary for resolving even moderately sized data sets. `taxadb` fills an important gap between existing tools and typical research patterns by providing a fast, reproducible approach for matching names to taxonomic identifiers. `taxadb` is not intended as an improvement or replacement for any existing approaches to taxonomic name resolution. In particular, `taxadb` is not a replacement for the APIs or databases provided, but merely an interface to taxonomic naming information contained within that data.

* cite other tools in the space (taxize + citations)
* Frandz citations - grumpy systemitists; we're not doing anything inferential
* no spell check/name modification
* could be more than one name for rank
* versioning paragraph - argument for version like provider; we process every so often
  * comment on the release schedule, rolling is not always up to date
